---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

# PristineSeasR2 

<!-- badges: start -->
[![R-CMD-check](https://github.com/pristine-seas/PristineSeasR2/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/pristine-seas/PristineSeasR2/actions/workflows/R-CMD-check.yaml)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

**PristineSeasR2** is the data science toolkit for [National Geographic Pristine Seas](https://www.nationalgeographic.org/society/our-programs/pristine-seas/). It provides standardized vocabularies, validation tools, color palettes, and utilities that ensure consistency across our marine conservation research.

## Installation
 
```r
# install.packages("pak")
pak::pak("pristine-seas/PristineSeasR2")
```

## Standardized Vocabularies

Controlled vocabularies ensure consistent data entry across expeditions and analysts:

```{r vocab-example}
library(PristineSeasR2)

# Five standardized vocabularies
names(allowed_vocab)

# Trophic classifications for reef fish
allowed_vocab$trophic_group
```

### Data Validation

Catch non-standard entries early — `validate_vocab()` stops your pipeline if invalid values are found:
 
```{r validate-success}
# Valid data passes
validate_vocab(c("shark", "top_predator", "planktivore"), "trophic_group")
```

```{r validate-error, error=TRUE}
# Invalid data stops with informative error
validate_vocab(c("shark", "apex_predator"), "trophic_group")
```

## Color Palettes

Publication-ready colors designed for marine science visualization:

```{r palettes, fig.height=5, fig.width=8, dpi=150}
library(ggplot2)

# Example: Trophic group colors
trophic_colors <- ps_colors("trophic_group")

# Sample data
fish_data <- data.frame(
  trophic_group = factor(names(trophic_colors), levels = names(trophic_colors)),
  biomass = c(45, 120, 280, 190, 85)
)

ggplot(fish_data, aes(x = trophic_group, y = biomass, fill = trophic_group)) +
  geom_col(width = 0.7) +
  scale_fill_ps("trophic_group") +
  labs(
    title = "Reef Fish Biomass by Trophic Group",
    x = NULL,
    y = expression(Biomass~(g/m^2))
  ) +
  theme_ps() +
  theme(legend.position = "none")
```

### Available Palettes

```{r show-palettes, echo=FALSE, fig.height=6, fig.width=9, dpi=150}
library(ggplot2)
library(dplyr)
library(tidyr)

# Build palette display data
palette_names <- c("trophic_group", "depth_strata", "functional_groups")

palette_data <- lapply(palette_names, function(pal) {
  colors <- ps_colors(pal)
  data.frame(
    palette = pal,
    value = factor(names(colors), levels = names(colors)),
    color = unname(colors),
    stringsAsFactors = FALSE
  )
}) |> bind_rows()

palette_data$palette <- factor(
  palette_data$palette, 
  levels = palette_names,
  labels = c("trophic_group", "depth_strata", "functional_groups")
)

ggplot(palette_data, aes(x = value, y = 1, fill = color)) +
 geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_identity() +
  facet_wrap(~palette, scales = "free_x", ncol = 1) +
  labs(title = "PristineSeasR2 Color Palettes") +
  theme_void() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5, margin = margin(b = 15)),
    strip.text = element_text(size = 11, face = "bold", margin = margin(b = 8)),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8, margin = margin(t = 5)),
    plot.margin = margin(20, 20, 20, 20)
  )
```

## Depth Stratification

Convert continuous depths to standardized survey categories:

```{r stratify}
depths_m <- c(2, 8, 12, 22, 45)

# Classify into strata
stratify(depths_m)

# Get station suffixes for IDs
station_suffix(stratify(depths_m))
```

## Key Functions
 
| Function | Purpose |
|----------|---------|
| `allowed_vocab` | Standardized vocabularies for survey data |
| `validate_vocab()` | Validate data against vocabularies |
| `ps_colors()` | Get color palettes |
| `scale_fill_ps()` / `scale_color_ps()` | ggplot2 color scales |
| `theme_ps()` / `theme_ps_map()` | ggplot2 themes |
| `stratify()` | Depth stratification |
| `get_drive_paths()` | Access shared Google Drive folders |
| `create_expedition()` | Create expedition folder structure |

## Learn More

- [Function Reference](https://pristine-seas.github.io/PristineSeasR2/reference/)
- [Pristine Seas Website](https://www.nationalgeographic.org/society/our-programs/pristine-seas/)

---

<p align="center">
<em>Developed by the Pristine Seas Science Team</em>
</p>
