---
title: "Thermal stress (DHW) from NOAA Coral Reef Watch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Thermal stress (DHW) from NOAA Coral Reef Watch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = TRUE,
  warning  = FALSE
)

library(ggplot2)
library(PristineSeasR2)
library(sf)
library(dplyr)
library(purrr)
```

This vignette shows how to retrieve **NOAA Coral Reef Watch Degree Heating Week (DHW)** (°C-weeks) from NOAA's ERDDAP servers using **PristineSeasR2**.
 
**DHW** is a cumulative measure of heat stress that integrates both intensity and duration of anomalous temperatures. It is widely used as a high-level indicator of bleaching risk and historical thermal exposure.

You can query DHW using:

- `get_crw_dhw_bbox()` for a numeric bounding box (lat/lon), or
- `get_crw_dhw_sf()` for an `sf` feature (the function derives a bbox from the geometry).

By default, `get_crw_dhw_sf()` returns a **daily mean DHW** across all grid cells within the feature's bounding box

- `summarise_daily = TRUE` **(default)** - daily mean DHW time series (recommended for comparisons)
- `summarise_daily = FALSE` - raw gridded values (one row per grid cell per day), useful for QC or spatial heterogeneity checks

> Practical note: these helpers query by **bounding box**. This is typically stable at the CRW grid resolution and is fast and robust for atoll or island scale comparisons.

# Query by coords (numeric bbox)

Use `get_crw_dhw_bbox()` when you already know your bounding box in WGS84 (EPSG:4326). This returns **gridded** values (lat/lon pixels) for each day.

```{r bbox-example}
# Bounding box for Jemo Island, Marshall Islands
lat_min <- 10.0547
lat_max <- 10.1306
lon_min <- 169.5001
lon_max <- 169.5905

dhw_grid <- get_crw_dhw_bbox(lat_min = lat_min,
                             lat_max = lat_max,
                             lon_min = lon_min,
                             lon_max = lon_max,
                             start   = "2023-01-01",
                             end     = "2025-12-31",
                             verbose = TRUE)

head(dhw_grid)
```

### Summarize by day (daily mean across the bbox)

A common pattern is to collapse the gridded output into a single daily mean DHW time series, which is what `get_crw_dhw_sf()` does by default.

```{r summarize-grid}
dhw_daily <- dhw_grid |>
  group_by(date) |>
  summarise(avg_dhw = mean(dhw, na.rm = TRUE), .groups = "drop")

head(dhw_daily)
```

# Query by `sf` feature (recommended)

Use `get_crw_dhw_sf()` when you have an `sf` polygon (atoll, reef area, MPA boundary, site footprint, etc.). The function will:

1) transform the geometry to WGS84 (EPSG:4326)
2) compute the bounding box
3) query ERDDAP in yearly chunks
4) **return a daily mean DHW time series by default** (`summarise_daily = TRUE`)

```{r create-aois}
# Create sf features for Samoa islands from bounding box coordinates

# Savaii
savaii <- st_as_sfc(st_bbox(c(xmin = -172.87, xmax = -172.12, 
                              ymin = -13.90, ymax = -13.31), 
                            crs = 4326)) |>
  st_as_sf() |>
  mutate(island = "Savaii")

# Upolu  
upolu <- st_as_sfc(st_bbox(c(xmin = -172.12, xmax = -171.34,
                             ymin = -14.16, ymax = -13.63),
                           crs = 4326)) |>
  st_as_sf() |>
  mutate(island = "Upolu")

# Combine
AOIs <- bind_rows(savaii, upolu)
AOIs

```

## Single AOI example

```{r single-aoi}
savaii_dhw <- get_crw_dhw_sf(feature  = savaii,
                             start    = "2023-01-01",
                             end      = "2024-12-31",
                             name_col = "island",
                             out_col  = "island",
                             summarise_daily = TRUE,
                             verbose  = TRUE)

savaii_dhw
```

# Multiple AOIs

For multiple features, iterate over rows and bind results. This keeps the core function simple (one feature in, one feature out) while still supporting batch workflows.

```{r multiple-aois}
dhw_samoa <- map_dfr(seq_len(nrow(AOIs)), \(i) {
  get_crw_dhw_sf(feature  = AOIs[i, ],
                 start    = "2023-01-01",
                 end      = "2025-12-31",
                 name_col = "island",
                 out_col  = "island",
                 summarise_daily = TRUE,
                 verbose  = TRUE)
})

dhw_samoa
```

# Plotting DHW time series

This plot visualizes the daily mean DHW histories for both islands. The horizontal reference lines are common heuristics:

- **4 DHW**: elevated bleaching risk
- **8 DHW**: severe bleaching / high mortality risk

These are rules of thumb and should be interpreted with context (seasonality, species composition, local adaptation, survey observations).

```{r plot-dhw, fig.width=8, fig.height=5}
x_annot <- min(dhw_samoa$date, na.rm = TRUE)

dhw_samoa |>
  ggplot(aes(x = date, y = avg_dhw, color = island)) +
  geom_line(linewidth = 0.9, alpha = 0.85) +
  geom_hline(yintercept = 4, linetype = "dashed", color = "grey30", linewidth = 0.5) +
  geom_hline(yintercept = 8, linetype = "dotted", color = "grey30", linewidth = 0.5) +
  annotate("text", x = x_annot, y = 4.5, label = "Bleaching threshold (~4 DHW)",
           size = 3, color = "grey30", hjust = 0) +
  annotate("text", x = x_annot, y = 8.5, label = "Severe threshold (~8 DHW)",
           size = 3, color = "grey30", hjust = 0) +
  scale_color_manual(values = c("Savaii" = "#2E6F95", "Upolu" = "#E07A5F")) +
  scale_x_date(date_breaks = "3 months", date_labels = "%Y-%m") +
  scale_y_continuous(name = "DHW (°C-weeks)", limits = c(0, NA)) +
  labs(title = "Thermal stress histories: Samoa",
       subtitle = "NOAA Coral Reef Watch Degree Heating Weeks (DHW)",
       x = NULL,
       color = NULL,
       caption = "Source: NOAA Coral Reef Watch via ERDDAP") +
  theme_ps()
```
